Setting parameters for biological models with ANIMO

### 2014-01-30 Paper.tex
Revising the paper. Added the true image for the parameter sweep result. Now working on a workflow figure.

\begin{abstract}
ANIMO (Analysis of Networks with Interactive MOdeling) is a software for modeling biological networks, such as e.g. signaling, metabolic or gene networks.
An ANIMO model is essentially the sum of a network topology and a number of interaction parameters. The topology describes the interactions between biological entities in form of a graph, while the parameters determine the speed of occurrence of such interactions.

When a mismatch is observed between the behavior of an ANIMO model and experimental data, we want to update the model so that it explains the new data. In general, the topology of a model can be expanded with new (known or hypothetical) players, and enable it to match experimental data. However, the unrestrained addition of new parts to a model causes two problems: models can become too complex too fast, to the point of being intractable, and too many parts marked as "hypothetical" or "not known" make a model unrealistic. Even if changing the topology is normally the easier task, these problems push us to try a better parameter fit as a first step, and resort to modifying the model topology only as a last resource.

@@

% - computational modeling in biology
% - assumption of robustness
% - ANIMO, piccolo esempio. ANIMO focuses on topology.
\section{Introduction}
The investigation of biological processes relies on computational support on a daily basis.
This happens not only because of the extremely large amount of data generated in the \emph{-omics era},
but also because many processes are simply too complex to be understood by the human brain alone.
For this reason, systems biology~\cite{priami-alg-sys-bio} has become more and more important in the last several years.

A single biological network such as a signaling or gene network, may involve up to hundreds of
different players. As it would be very difficult to understand the dynamic behavior of such
networks just by looking at their static representations, many tools were built~\cite{biopepa-interface,
blenx, cell-illustrator, copasi, ginsim, gna} to help the biologists define models of \emph{executable biology}~\cite{ex-bio}.
ANIMO (Analysis of Networks with Interactive MOdeling~\cite{animo-ieee}) is one of such tools. Its primary objective is to let the expert biologists
work directly on the formalization of their knowledge, supporting the generation of new insights on the studied processes~\cite{animo-gene}.
An ANIMO model is formed by two main parts: the network topology and the parameters.
The topology describes which biological components are included in the model, and which are the
interactions we want to represent. The parameters define the rate of occurrence of such interactions,
which are described based on simplified kinetic formulae.

Proteins are normally expressed at different concentrations in different individuals of the same species, and yet the
overall behavior of their biological networks does not differ significantly. This phenomenon has led to the notion that biological
networks are inherently \emph{robust}~\cite{bio-robust1, bio-robust2}. In modeling terms, this means that,
if a model is a close representation of a biological network, most of the parameters of that model
can vary inside a certain interval without influencing the qualitative behavior of the whole network.
The interactive approach of ANIMO is based on the assumption of robustness, as our tool is mainly aimed at the
development of network models with a focus on the topology.
Ideally, the biologist can ``play'' with the topology of a network, working
more towards matching the qualitative behavior of experimental data, rather than precisely reproducing it.
However, it is not our intention to concentrate exclusively on the network topology:
in many cases a better parameter choice can improve the behavior of a network more than the addition of new components.
Indeed, making an unnecessarily complex model could reduce its usefulness both in terms of analysis
performances and closeness to reality. The first problem is simply due to the complexity of a network,
which would require more and more computational resources to be analysed\footnote{We also refer to the
problem of \emph{state space explosion}: when a model contains too many loosely coupled components,
the set of its possible evolutions grows exponentially, to the point of making it impossible to apply some analysis
techniques.}.
Realism of network models is more related to their
ultimate usefulness: a model that explains a particular behavior very well but contains many nodes marked
as ``unknown'' has little applicability, as its connection with known processes is very loose.
Therefore, it is desirable that a better parameter set is regularly sought for during the design cycle of
a complex biological network model. Some support for parameter choice was already provided in the first versions of ANIMO.
We present here an extended set of tools aimed at achieving a closer fit between ANIMO models and experimental data.
A guideline on how to use these tools to get the best results will also be presented as an ideal workflow.
Thanks to the better awareness on parameter choice gained through this new extension of ANIMO, the biologists
will be able to judge more easily which are the most promising topologies for a network, and thus drive
the experimental research more efficiently.

% - topology - parameters
% - problemi: small example network with parameter variations (also: some parts of a network are more sensitive to parameter variations than others)
\section{ANIMO models}
The starting point of ANIMO is the traditional static representation of biological networks,
which can be easily drawn and managed in softwares like Cytoscape~\cite{cytoscape}. Indeed, ANIMO was implemented
as a plug-in to Cytoscape, with the aim of adding dynamics to the static representation of biological
networks, and thus allow for analysis on the behavior of such networks. The user interface of ANIMO
can be seen in Figure~\ref{fig:animo-ui}: at the center is the Network panel, where the network
model is represented in the familiar nodes-edges form used in the domain of biology.
Models in ANIMO are \emph{activity-based}, in the sense that nodes have their activity as main
property, and interactions among nodes change the activity of their targets. The concept of activity
can be intended for example as a generic post-translational modification a molecule can undergo to change its
function. In the case of a kinase, the phosphorylated state is usually interpreted as active.
In the context of a gene network, the activity of a node standing for a gene represents its current
transcriptional status. A basic type of analysis that can be performed with ANIMO is the generation of
a simulation run, which is presented to the user in the form of a graph (to the right). The activity
graphs generated by ANIMO show the variation in activity of selected nodes over the course of the simulation run.
In addition to that, a slider placed under each graph allows the user to color the nodes in the Network panel
depending on their activity level at any point during the course of the simulation: the Legend
panel on the left links colors to activity. ANIMO models are based on the formalism of \tas~\cite{timed-automata-alur-dill},
which are not directly shown to the user. Indeed, ANIMO was designed with the aim of enabling expert biologists
to formalize their knowledge without the need for additional mathematical training.

\begin{figure}
\begin{center}
  \includegraphics[width=.6\textwidth]{images/animo-ui}
  \caption{ANIMO user interface.\label{fig:animo-ui}}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
  \includegraphics[width=.4\textwidth]{images/animo-edge-ui}
  \caption{Editing the parameters for an ANIMO interaction.\label{fig:animo-edge-ui}}
\end{center}
\end{figure}

Of the two main components of an ANIMO model, only the topology is immediately visible to the user in the
Network panel; the parameters are accessed by double clicking the arcs representing node-node interactions.
The dialog window that is shown for an interaction contains the details of the abstract reaction kinetic
describing the interaction, together with the current value of its parameter $k$ (see Fig.~\ref{fig:animo-edge-ui}).
The unique parameter associated to any interaction in ANIMO is used as a scale factor to make the modelled
reaction occur faster or slower. We also provide the user with pre-set values for
$k$, encouraging an initial qualitative assignment of reaction rates as ``slow'', ``fast'' and so forth.
This approach is based on the assumption that biological networks are inherently robust:
once an acceptable set of parameters is found, a more precise parameter search will generally have
little impact on the fitness of a model to a reference data set.
However, as can be seen in the small example in Figure~\ref{fig:small-example-feedback},
robustness does not imply that any parameter choice will do. In particular, network motifs
such as feedback loops make models more dependent on parameter settings.
As biological networks tend to heavily rely on cross-talk, network topologies can rapidly become
complex. Some manual parameter fitting is currently necessary for the more complex ANIMO models.
This work is sometimes slow and error-prone, taking away some of the user-friendliness
for which ANIMO aims.
We will show how ANIMO has been improved to make parameter choice easier for the user.

@@

The scenarios are used to choose which are the nodes to be taken into account
when computing the speed at which an interaction occurs.
\begin{description}
  \item[Scenario 1] the simplest of the three, this scenario approximates an interaction
    without taking into account the abundance of substrate: the activation/inhibition rate
    depends only on the activity level of the upstream node. In the MEK $\rightarrow$ ERK
    example, the rate of the interaction would be linearly dependent on the current activity level of MEK.
  \item[Scenario 2] differently from scenario 1, scenario 2 models the interaction rate
    to be directly dependent both on the activity
    level of the upstream node and on the availability of substrate. By substrate we
    mean the inactive downstream node in case of activation interaction, and active downstream node in case of inhibition.
    In the above example, the interaction rate is linearly dependent on
    both the activity level of MEK and the inactive fraction of ERK.
    The inactive level of a node is computed by subtracting the current activity level from the maximum activity.
  \item[Scenario 3] the nodes on which this scenario depends can be chosen directly by the user,
    and can be either the active or inactive fraction of any two nodes in the network.
    This scenario can be used to represent an AND-gate, where two nodes are required to
    be simultaneously active in order to influence their target.
\end{description}

The choice for the parameter $k$ can be made directly by inserting a numeric value, or 
indirectly by choosing a preset among the proposed qualitative values \emph{very slow, slow, medium, fast, very fast}.
We advise our users to base their initial parameter choice on these qualitative presets.
In Figure~\ref{fig:small-example-feedback}, all parameters are initially set to \emph{medium}
(Fig.~\ref{fig:small-example-feedback}{\sf a}).
The peak in EGFR was obtained setting the value of $k$ for {\sf EGFR internalization}~$\dashv$~{\sf EGFR}
to \emph{fast} (Fig.~\ref{fig:small-example-feedback}{\sf b}). As this is not enough
to get a peak also for ERK, which remains constantly inactive, a lower value of $k$
for {\sf ERK phosphatase}~$\dashv$~{\sf ERK} can be tried. A first attempt with
\emph{slow} (Fig.~\ref{fig:small-example-feedback}{\sf c}) leads to a low peak,
which can be increased by further lowering that parameter to \emph{very slow}
(Fig.~\ref{fig:small-example-feedback}{\sf d}). Further adjustment of the parameters
and scenario choices other than 1 leads to the graph shown in Figure~\ref{fig:animo-ui}.

@@

% Concludi con esempio di sweep sull'esempio abcd (erk etc): son 3125 simulazioni se lo fai
%  andare solo sui parametri ``slow'', ``medium'', etc e con le 5 interazioni di abcde
We now present an example application of ANIMO's parameter sweep feature on the model shown in Figure~\ref{fig:animo-ui}.
A logarithmic-scale was selected, letting the parameter values
vary over the 5 pre-set values ``very slow'' ($k = 0.001$), ``slow'' ($k = 0.002$), \dots, ``very fast'' ($k = 0.016$)
for all 5 interactions in the network. This resulted in $5 ^ 5 = 3125$ simulations, that were performed
in 34 seconds on an eight-core Intel\circledR\ Core$^{\mbox{\scriptsize\texttrademark}}$ i7 CPU at 2.80GHz equipped with 4 Gb RAM
and running Ubuntu GNU/Linux 13.10 64bit. The result window shown by ANIMO at the end of the computation can
be seen in Figure~\ref{fig:parameter-sweep-result}, where the best 6 results are listed in form of graphs.
By interacting with the interface, the user can choose to see more of the top-scoring results: this will allow to check
the width of the parameter settings for which the wanted behavior can still be considered ``close enough'' to
the experimental data.

\begin{figure}
\centering
\begin{picture}(350,260)
\put(0, 0){\includegraphics[width=0.8\textwidth]{images/parameter_sweep_results}}
%\put(150,50){\hbox{Over here}}
\put(70, 70){\includegraphics[width=0.5\textwidth]{images/dummy_sign}}
\end{picture}
\caption{Result window of a logarithmic parameter sweep in the ANIMO model of Figure~\ref{fig:animo-ui}.\label{fig:parameter-sweep-result}}
\end{figure}

>>>>

\begin{abstract}
ANIMO (Analysis of Networks with Interactive MOdeling) is a software for modeling biological networks, such as e.g. signaling, metabolic or gene networks.
An ANIMO model is essentially the sum of a network topology (describing reactants and interaction types in form of a graph) and a number of interaction parameters. The topology describes the interactions between biological entities in form of a graph, while the parameters determine the speed of occurrence of such interactions.

When a mismatch is observed between the behavior of an ANIMO model and experimental data, we want to update the model so that it explains the new data. In general, the topology of a model can be expanded with new (known or hypothetical) nodes, and enables it to match experimental data. However, the unrestrained addition of new parts to a model causes two problems: models can become too complex too fast, to the point of being intractable, and too many parts marked as "hypothetical" or "not known" make a model unrealistic. Even if changing the topology is normally the easier task, these problems push us to try a better parameter fit as a first step, and resort to modifying the model topology only as a last resource.

@@

% - computational modeling in biology
% - assumption of robustness
% - ANIMO, piccolo esempio. ANIMO focuses on topology.
\section{Introduction}
The investigation of biological processes relies on computational support on a daily basis.
This happens not only because of the extremely large amount of data generated in the \emph{-omics era},
but also because many processes are simply too complex to be understood without appropriate modeling tools.
For this reason, systems biology~\cite{priami-alg-sys-bio} has become more and more important in the last several years.

A single biological network such as a signaling or gene network, may involve up to hundreds of
different molecular species. As it would be very difficult to understand the dynamic behavior of such
networks just by looking at their static representations, many tools were built~\cite{biopepa-interface,
blenx, cell-illustrator, copasi, ginsim, gna} to help the biologists define models of \emph{executable biology}~\cite{ex-bio}.
ANIMO (Analysis of Networks with Interactive MOdeling~\cite{animo-ieee}) is one of such tools. Its primary objective is to let the expert biologists
work directly on the formalization of their knowledge, supporting the generation of new insights on the studied processes~\cite{animo-gene}.
An ANIMO model is formed by two main parts: the network topology and the parameters.
The topology describes which biological components are included in the model, and which are the
interactions we want to represent. The parameters define the rate of occurrence of such interactions,
which are described based on simplified kinetic formulae.

Proteins are normally expressed at different concentrations in different individuals of the same species, and yet the
overall behavior of their biological networks does not differ significantly. This phenomenon has led to the notion that biological
networks are inherently \emph{robust}~\cite{bio-robust1, bio-robust2}. In modeling terms, this means that,
if a model is a close representation of a biological network, most of the parameters of that model
can vary inside a certain interval without influencing the qualitative behavior of the whole network.
The interactive approach of ANIMO is based on the assumption of robustness, as our tool is mainly aimed at the
development of network models with a focus on the topology.
Ideally, the biologist can ``play'' with the topology of a network, working
more towards matching the qualitative behavior of experimental data, rather than precisely reproducing it.
However, it is not our intention to concentrate exclusively on the network topology:
in many cases a better parameter choice can improve the behavior of a network more than the addition of new components.
Indeed, making an unnecessarily complex model could reduce its usefulness both in terms of analysis
performances and closeness to reality. The first problem is simply due to the complexity of a network,
which would require more and more computational resources to be analysed\footnote{We also refer to the
problem of \emph{state space explosion}: when a model contains too many loosely coupled components,
the set of its possible evolutions grows exponentially, to the point of making it impossible to effectively apply analysis
techniques.}.
Realism of network models is more related to their
ultimate usefulness: a model that explains a particular behavior very well but contains many nodes marked
as ``unknown'' has little applicability, as its connection with known processes is very loose.
Therefore, it is desirable that a better parameter set is regularly sought for during the design cycle of
a complex biological network model. Some support for parameter choice was already provided in the first versions of ANIMO.
We present here an extended set of tools aimed at achieving a closer fit between ANIMO models and experimental data.
A guideline on how to use these tools to get the best results will also be presented as an ideal workflow.
Thanks to the better awareness on parameter choice gained through this new extension of ANIMO, the biologists
will be able to judge more easily which are the most promising topologies for a network, and thus drive
the experimental research more efficiently.

% - topology - parameters
% - problemi: small example network with parameter variations (also: some parts of a network are more sensitive to parameter variations than others)
\section{ANIMO models}
The starting point of ANIMO is the traditional static representation of biological networks,
which can be easily drawn and managed in softwares like Cytoscape~\cite{cytoscape}. Indeed, ANIMO was implemented
as a plug-in to Cytoscape, with the aim of adding dynamics to the static representation of biological
networks, and thus allow for analysis on the behavior of such networks. The user interface of ANIMO
can be seen in Figure~\ref{fig:animo-ui}: at the center is the Network panel, where the network
model is represented in the familiar nodes-edges form used in the domain of biology.
Models in ANIMO are \emph{activity-based}, in the sense that nodes have their activity as main
property, and interactions among nodes change the activity of their targets. The concept of activity
can be intended for example as a generic post-translational modification a molecule can undergo to change its
function. In the case of a kinase, the phosphorylated state is usually interpreted as active.
In the context of a gene network, the activity of a node standing for a gene represents its current
transcriptional status. A basic type of analysis that can be performed with ANIMO is the generation of
a simulation run, which is presented to the user in the form of a graph (to the right). The activity
graphs generated by ANIMO show the variation in activity of selected nodes over the course of the simulation run.
In addition to that, a slider placed under each graph allows the user to color the nodes in the Network panel
depending on their activity level at any point during the course of the simulation: the Legend
panel on the left links colors to activity. ANIMO models are based on the formalism of \tas~\cite{timed-automata-alur-dill},
which is not directly shown to the user. Indeed, ANIMO was designed with the aim of enabling expert biologists
to formalize their knowledge without the need for additional mathematical training.

\begin{figure}[htbp]
\begin{center}
  \includegraphics[width=.59\textwidth]{images/animo-ui}
  \caption{ANIMO user interface.\label{fig:animo-ui}}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
  \includegraphics[width=.4\textwidth]{images/animo-edge-ui}
  \caption{Editing the parameters for an ANIMO interaction.\label{fig:animo-edge-ui}}
\end{center}
\end{figure}

Of the two main components of an ANIMO model, only the topology is immediately visible to the user in the
Network panel; the parameters are accessed by double clicking the arcs representing node-node interactions.
The dialog window that is shown for an interaction contains the details of the abstract reaction kinetic
describing the interaction, together with the current value of its parameter $k$ (see Fig.~\ref{fig:animo-edge-ui}).
The unique parameter associated to any interaction in ANIMO is used as a scale factor to make the modelled
reaction occur faster or slower. We also provide the user with pre-set values for
$k$, encouraging an initial qualitative assignment of reaction rates as ``slow'', ``fast'' and so forth.
This approach is based on the assumption that biological networks are inherently robust:
once an acceptable set of parameters is found, a more precise parameter search will generally have
little impact on the fitness of a model to a reference data set.
However, as can be seen in the small example in Figure~\ref{fig:small-example-feedback},
robustness does not imply that any parameter choice will do.
In that example, all parameters are initially set to \emph{medium}
(Fig.~\ref{fig:small-example-feedback}{\sf a}).
The peak in EGFR was obtained setting the value of $k$ for {\sf EGFR internalization}~$\dashv$~{\sf EGFR}
to \emph{fast} (Fig.~\ref{fig:small-example-feedback}{\sf b}). As this is not enough
to get a peak also for ERK, which remains constantly inactive, a lower value of $k$
for {\sf ERK phosphatase}~$\dashv$~{\sf ERK} can be tried. A first attempt with
\emph{slow} (Fig.~\ref{fig:small-example-feedback}{\sf c}) leads to a low peak,
which can be increased by further lowering that parameter to \emph{very slow}
(Fig.~\ref{fig:small-example-feedback}{\sf d}). Further adjustment of the parameters
and non-default scenario choices leads to the graph shown in Figure~\ref{fig:animo-ui}.
Network motifs such as the feedback loop shown in Figure~\ref{fig:small-example-feedback}
make models more dependent on parameter settings.
As biological networks tend to heavily rely on cross-talk, network topologies can rapidly become
complex. Some manual parameter fitting is currently necessary for the more complex ANIMO models.
This work is sometimes slow and error-prone, taking away some of the user-friendliness
for which ANIMO aims.
We will show how ANIMO has been improved to make parameter choice easier for the user.

@@

The scenarios are used to choose which are the nodes to be taken into account
when computing the speed at which an interaction occurs.
\begin{description}
  \item[Scenario 1] the simplest of the three, this scenario approximates an interaction
    without taking into account the abundance of substrate: the activation/inhibition rate
    depends only on the activity level of the upstream node. In the MEK $\rightarrow$ ERK
    example, the rate of the interaction would be linearly dependent on the current activity level of MEK.
  \item[Scenario 2] scenario 2 models the interaction rate
    to be directly dependent both on the activity
    level of the upstream node and on the availability of substrate. By substrate we
    mean the inactive downstream node in case of activation interaction, and active downstream node in case of inhibition.
    In the above example, the interaction rate is linearly dependent on
    both the activity level of MEK and the inactive fraction of ERK.
    The inactive level of a node is computed by subtracting the current activity level from the maximum activity.
  \item[Scenario 3] the nodes on which this scenario depends can be chosen directly by the user,
    and can be either the active or inactive fraction of any two nodes in the network.
    This scenario can be used to represent an AND-gate, where two nodes are required to
    be simultaneously active in order to influence their target.
\end{description}

The choice for the parameter $k$ can be made directly by inserting a numeric value, or 
indirectly by choosing a preset among the proposed qualitative values \emph{very slow, slow, medium, fast, very fast}.
% We advise our users to base their initial parameter choice on these qualitative presets.

@@

% Concludi con esempio di sweep sull'esempio abcd (erk etc): son 3125 simulazioni se lo fai
%  andare solo sui parametri ``slow'', ``medium'', etc e con le 5 interazioni di abcde
We now present an example application of ANIMO's parameter sweep feature on the model shown in Figure~\ref{fig:animo-ui}.
This allows us to make a more exhaustive search than the one presented in Figure~\ref{fig:small-example-feedback},
using an automatic approach instead of a manual search.
A logarithmic parameter sweep was performed, letting the parameter values
vary over the 5 pre-sets ``very slow'' ($k = 0.001$), ``slow'' ($k = 0.002$), \dots, ``very fast'' ($k = 0.016$)
for all 5 interactions in the network. This resulted in $\mbox{\it parameter choices}^{\mbox{\it\scriptsize interactions}} = 5 ^ 5 = 3\,125$ simulations, that were computed
in 43 seconds %43 s with 16 threads. It is 57 with 8 threads (cpu is exploited less than 100%)
on an eight-core Intel\circledR\ Core$^{\mbox{\scriptsize\texttrademark}}$ i7 CPU at 2.80GHz equipped with 4 Gb RAM
and running Ubuntu GNU/Linux 13.10 64bit. The model we consider here is small enough for one simulation to take
less than a second, but the exponential growth of the number of simulations required for a parameter sweep
makes the sheer number of tasks challenging for any processing unit.
% Indeed, a more thorough
% parameter sweep that considers 10 different setting for each of the 5 interactions of the considered
% model would require the computation of $10^5 = 100\,000$ simulations.
Indeed, a more complex model involving 10 interactions instead of 5 would require $5^{10} = 9\,765\,625$ simulations,
which could be computed in approximately a day and a half on an eight-core machine. Yet, such an
estimation would still be better than the twelve days it would take if the simulations were to be computed in sequence.
% the same experiment
% was run executing only one simulation at a time, and it took about 7 minutes and 34 seconds, %453837 ms
% which is in line with the fact that this type of analysis is trivially parallelizable.

The result window shown by ANIMO at the end of the computation of the $3\,125$ simulations can
be seen in Figure~\ref{fig:parameter-sweep-result}, where the best 6 results are listed in form of graphs.
By interacting with the interface, the user can choose to see more of the top-scoring results: this will allow to check
the width of the parameter settings for which the wanted behavior can still be considered ``close enough'' to
the experimental data. Pressing one of the buttons labelled ``I want this'' under one of the graphs
allows the user to copy the parameters used in the selected simulation to the network model.

\begin{figure}
\centering
\includegraphics[width=0.92\textwidth]{images/parameter_sweep_results}
%\begin{picture}(350,260)
%\put(0, 0){\includegraphics[width=0.8\textwidth]{images/parameter_sweep_results}}
%\put(150,50){\hbox{Over here}}
%\put(70, 70){\includegraphics[width=0.5\textwidth]{images/dummy_sign}}
%\end{picture}
\caption{Result window of a logarithmic parameter sweep in the ANIMO model of Figure~\ref{fig:animo-ui}.\label{fig:parameter-sweep-result}}
\end{figure}
